### 2021-10-21 submission/manuscript.tex
Address reviewer's comments

@@

\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex

@@

  \usepackage{amssymb}

@@

\usepackage{longtable,booktabs}

@@

@@

\ifluatex

@@

\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}

@@

Numerous studies have attempted to address the problem of removing or
``denoising'' artifacts from data generated using Illumina's MiSeq
platform. In one approach, paired sequence reads are aligned and any
discrepencies between the reads are resolved based on the difference in
quality score for the position in question (2--4); quality scores are
also used to curate single reads (5). In addition, a polishing step is
often used to identify ASVs based on the frequency and similarity of
sequences (2, 3, 6). In a second approach, the quality scores and types
of errors are modelled to cluster sequence reads directly into ASVs (7).
Regardless of the approach, many pipelines advocate for abundance-based
screening where rare sequences are removed from each dataset prior to
outputting the sequence data as ASVs (3, 6, 7). Some algorithms
recommend removing all ASVs that appear one (i.e.~singletons) (7), eight
(3), or ten (6) or fewer times; these pipelines also vary in whether the
minimum abundance threshold should be applied to individual samples (6,
7) or the pool of samples in a study (3). A notable exception, the
mothur-based pipeline discourages the practice of removing rare
sequences (2). After assigning sequences to ASVs, ASVs are often
analyzed as a taxonomic unit or clustered to generate operational
taxonomic units (OTUs) or phylotypes.

@@

\textbf{Datasets.} I collected 12 publicly available datasets that used
the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
gene from a variety of environments (Table 1; Figure S1). After removing
poor quality and chimeric ASVs and samples that had uncharacteristically
low number of sequences for the dataset, these datasets included between
7 and 490 samples. The median number of sequences for each dataset
ranged between 6,477 and 193,464. Strikingly, aside from the relatively
small marine and soil datasets, the difference between the sample with
the fewest sequences and the sample with the most sequences for each
dataset varied by between 7.4 and 96.6-fold.

\textbf{The nature of singletons.} Removal of rare ASVs is commonly
justified as a method of removing ASVs that are artifacts. If such ASVs
are artifacts, then one would expect the number of singleton ASVs to
accumulate with sequencing depth. Contrary to this expectation, the
median percentage of sequences that were discarded when singleton ASVs
were removed from each dataset varied between 0.42 and 22.20\%
(bioethanol and seagrass). In addition, with the exception of the
samples from the marine and sediment datasets (Spearman correlation,
P\textgreater0.05), the fraction of singleton ASVs in samples was
negatively correlated with the number of sequences in each sample with a
range between -0.26 and -0.87 (rice and bioethanol) (Figures 1A and S2).
This showed that with additional sequencing, the probability of seeing
singleton ASVs in multiple samples was greater than the probability of
generating an artifact. This suggests that the singleton ASVs are not as
likely to be artifacts as previously thought. Furthermore, if singleton
ASVs were artifacts, then one would not expect to find them in other
samples from the same dataset. In fact, singleton ASVs from samples with
fewer sequences were often found in samples with more sequences. At
least 50\% of the singleton ASVs found in the samples from the mice,
rice, seagrass, and stream datasets were found in another sample from
the same dataset (Figure 1B). Considering the likelihood of finding an
ASV duplicated in another sample is confounded by the number of samples
and inter-sample diversity, the high coverage of singleton ASVs in these
datasets was remarkable. The correlation between the number of sequences
in a sample and the fraction of that sample's singleton ASVs that were
covered by another sample in the dataset was significant and negative
for 9 of the datasets ranging between -0.33 and -0.84 for the rice and
seagrass datasets, respectively ((Figures 1C and S3)). The negative
correlation indicated that the singleton ASVs in the smaller samples
were more likely to be covered by ASVs in the larger samples. Among the
three datasets without a significant correlation (Spearman correlation,
P\textgreater0.05), the marine and soil datasets had the fewest samples
in our collection and the stream dataset already had a high level of
coverage regardless of the number of sequences. Contrary to the common
motivation for removing rare ASVs, these results indicate that this
practice disproportionately impacts samples with fewer sequences and
likely removes more non-artifact ASVs than those that are artifacts.

@@

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
communities were randomly sampled from the null distribution as
described above. For the second treatment group, I randomly selected
10\% of the ASVs in the pooled study distribution to increase their
abundance by 5\%. I randomly generated 100 simulated sets of treatment
groups and samples. I then tested the ability to detect a difference
between the two treatment groups using alpha and beta diversity metrics.
The fraction of significant tests was a measurement of the statistical
power to detect the difference between the treatment groups. When
considering the differences in richness and diversity, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7).
Among the remaining datasets, the power to detect a difference in the
richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
between 0.10 and 0.53 (rainforest and stream) to detect a difference in
diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
removed, the power to detect a difference in the richness of ASVs
dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
40.0 and 93.3\% for their diversity(rainforest and soil; Figure 4B). The
effect of removing rare ASVs on the richness of OTUs and their diversity
was similar (Figure S7AB). I used the Bray-Curtis dissimilarity index to
compare the simulated communities within each dataset and calculated the
power to detect differences between the two simulated treatment groups
using the analysis of molecular variance (Figure 4C and S7C). Without
removing rare sequences, the power to detect a difference between the
two simulated treatment groups varied between 0.41 and 1.00 (rainforest
and rainforest). Aside from the bioethanol, human, and mice datasets,
the power to detect differences dropped by between 6.5 and 64.0\% (soil
and rice) when singletons were removed. However, when ASVs that occurred
10 or fewer times were removed from each sample, the power to detect
differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
results were observed when ASVs were clustered into OTUs. Removing rare
ASVs reduced the ability to detect simulated treatment effects using
metrics commonly used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S8A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S8A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups (9). When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S8B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S8B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

@@

The practice of removing rare sequences from samples seems to be a
response to researchers prioritizing the number of reads and length of
sequences over their quality (5). Previous work has shown that assembly
of fully overlapping sequence reads results in the lowest sequencing
error rates (2). The studies highlighted in this analysis sequenced the
V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
platform with their version 2 chemistry. The resulting data consists of
two 250 nt reads that span a region that is about 253 nt long. In
contrast, there has been a movement to sequence longer regions with
similar chemistry resulting in less overlap between the sequencing reads
(10, 11). Alternatively, others have prioritized increasing the number
of sequences per sample by sequencing the V4 region, but with paired or
single reads that are 150 nt long (12, 13). Both practices result in a
significantly higher error rate for the resulting assmebly. Instead,
researchers should prioritize the quality over the quantity and length
of their data. For these reasons, this analysis did not use lower
quality data generated by alternative methods.

The impacts of removing rare sequences on the representation of
communities are caused by the uneven impact of applying a single
abundance threshold across all samples. The number of sequences per
sample in the datasets highlighted in this analysis varied by as much as
100-fold (Table 1; Figure S1). In practice, I suspect this range is
actually larger since researchers may have opted against depositing
samples with fewer reads into the SRA. If the range in read coverage for
a study is 100-fold, a singleton in a small sample would have had a
comparable relative abundance as a sequence that appeared 100 times in a
more densely sequenced sample. However, it would have been removed from
the smaller sample and not the larger sample. Thus, applying a single
abundance threshold disproportionately impacted the samples with fewer
sequences. This seems to contradict the purpose of removing rare
sequences since a singleton in the smaller sample is more reliable than
the singleton in the larger sample. A superior approach to removing rare
sequences is to use rarefaction to conrol for uneven sampling.

Even with the best sequencing approaches, PCR artifacts and sequencing
errors persist. This may account for some sequences not being observed
other samples. Although this lack of inter-sample coverage could have
been due to treatment effects and natural variation, it is likely that
some portion of the sequences that were unique to samples were artifacts
or contained errors. Researchers are encouraged to use rarefaction to
control for uneven sampling and the presence of spurious sequences.
Previous work sequencing mock communities has shown that the number of
spurious sequences increases with sequencing depth (2, 14). By rarefying
data to a common number of sequences per sample, the number of spurious
sequences can be controlled. As shown in the data I presented, which was
rarefied to a common number of reads per sample within a dataset, when
rare sequences were not removed the power to detect differences was the
highest and the false discovery rate was the expected 5\% (Figures 5 and
S8).

In addition to considerations of how to control for the presence of
spurious sequences, researchers also need to be mindful of how to
interpret the results of their work. Because every dataset will contain
residual sequences that are spurious, measures of richness and diversity
should be made on a relative basis. For example, pronouncements that
communities from an environment or treatment contain a specific number
of taxa are problematic. Instead, we should limit ourselves to
indicating that samples from one treatment group has more taxa than
another without using absolute values of richness. Furthermore, we must
take caution in interpreting rare taxa. Although the data from the
studies highlighted here suggest that most rare sequences are not
spurious, it is likely that some are. Therefore, researchers must
approach rare sequences with more skepticism than more abundant
sequences. Reseachers should seek out other methods to confirm
inferences that they make about rare sequences. This is a standard that
should be applied regardless of their abundance (15).

How to curate and interpret rare sequences has been a significant
challenge since microbial ecologists transitioned away from Sanger
sequencing of samples (16--18). Although the extent of the ``rare
biosphere'' is still an open question, it is important to appreciate the
importance of rare populations in all communities. Populations can be
numerically rare but ubuiquitous or abundant and limited in their
geographic range. Alternatively, they can be numerically rare but
temporally common or abundant but present infrequently. Removing
sequences from any of these settings will limit our ability to study the
role of such populations or the processes that drive their patchy
distributions that are so common with microbial communities (8).

@@

\textbf{Data curation and analysis.} To insure the highest possible data
quality, datasets were limited to those where the 500 cycle version 2
MiSeq chemistry was used to sequence the amplicons. The paired 250 nt
reads resulted in near complete 2-fold sequencing coverage of every
nucleotide in the ca. 253 nt-long region. This region and sequencing
platform were selected because previous work has shown that a standard
data analysis pipeline in mothur results in a sequencing error rate
below 0.02\% (2). All sequence data were obtained from the Sequence Read
Archive (SRA) and processed using a standard mothur-based sequencing
pipeline that resulted in ASVs as generated by the pre.cluster algorithm
using a threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a
3\% distance threshold using mothur's cluster function with the
OptiClust algorithm (20). To minimize the effects of uneven sampling
effort, samples were rarefied to the number of sequences in the smallest
sample for each dataset. Because metrics of alpha diversity did not
consistently follow a normal distribution, I used the non-parametric
Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis
distances, the amova function within mothur was used, which implements
the analysis of molecular variance algorithm (21).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Knight2018}{}%
\CSLLeftMargin{1. }
\CSLRightInline{\textbf{Knight R}, \textbf{Vrbanac A}, \textbf{Taylor
BC}, \textbf{Aksenov A}, \textbf{Callewaert C}, \textbf{Debelius J},
\textbf{Gonzalez A}, \textbf{Kosciolek T}, \textbf{McCall L-I},
\textbf{McDonald D}, \textbf{Melnik AV}, \textbf{Morton JT},
\textbf{Navas J}, \textbf{Quinn RA}, \textbf{Sanders JG},
\textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Tripathi A},
\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhu Q}, \textbf{Caporaso
JG}, \textbf{Dorrestein PC}. 2018. Best practices for analysing
microbiomes. Nature Reviews Microbiology \textbf{16}:410--422.
doi:\href{https://doi.org/10.1038/s41579-018-0029-9}{10.1038/s41579-018-0029-9}.}

\leavevmode\hypertarget{ref-Kozich2013}{}%
\CSLLeftMargin{2. }
\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
a dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform}.
Applied and environmental microbiology \textbf{79}:5112--5120.}

\leavevmode\hypertarget{ref-Edgar2016}{}%
\CSLLeftMargin{3. }
\CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
error-correction for illumina 16S and {ITS} amplicon sequencing.
bio{R}xiv. doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.}

\leavevmode\hypertarget{ref-Masella2012}{}%
\CSLLeftMargin{4. }
\CSLRightInline{\textbf{Masella AP}, \textbf{Bartram AK},
\textbf{Truszkowski JM}, \textbf{Brown DG}, \textbf{Neufeld JD}. 2012.
{PANDAseq}: Paired-end assembler for illumina sequences. {BMC}
Bioinformatics \textbf{13}:31.
doi:\href{https://doi.org/10.1186/1471-2105-13-31}{10.1186/1471-2105-13-31}.}

\leavevmode\hypertarget{ref-Bokulich2013}{}%
\CSLLeftMargin{5. }
\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
vastly improves diversity estimates from illumina amplicon sequencing.
Nature Methods \textbf{10}:57--59.
doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}

\leavevmode\hypertarget{ref-Amir2017}{}%
\CSLLeftMargin{6. }
\CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
\textbf{Navas-Molina JA}, \textbf{Kopylova E}, \textbf{Morton JT},
\textbf{Xu ZZ}, \textbf{Kightley EP}, \textbf{Thompson LR}, \textbf{Hyde
ER}, \textbf{Gonzalez A}, \textbf{Knight R}. 2017. Deblur rapidly
resolves single-nucleotide community sequence patterns. {mSystems}
\textbf{2}.
doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.}

\leavevmode\hypertarget{ref-Callahan2016}{}%
\CSLLeftMargin{7. }
\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
SP}. 2016. {DADA}2: High-resolution sample inference from illumina
amplicon data. Nature Methods \textbf{13}:581--583.
doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}

>>>>

\usepackage{amsmath,amssymb}

@@

\usepackage{iftex}
\ifPDFTeX

@@

@@

\usepackage{longtable,booktabs,array}

@@

\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \setlength{\parskip}{#2\cslentryspacingunit}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}

@@

\ifLuaTeX

@@

@@

Numerous studies have attempted to address the problem of removing or
``denoising'' artifacts from data generated using Illumina's MiSeq
platform. In one approach, paired sequence reads are aligned and any
discrepencies between the reads are resolved based on the difference in
quality score for the position in question (2--4); quality scores are
also used to curate single reads (5). In addition, a polishing step is
often used to identify ASVs based on the frequency and similarity of
sequences (2, 3, 6). In a second approach, the quality scores and types
of errors are modelled to cluster sequence reads directly into ASVs (7).
Regardless of the approach, many pipelines advocate for abundance-based
screening where rare sequences are removed from each dataset prior to
outputting the sequence data as ASVs (3, 6, 7). Some algorithms
recommend removing all ASVs that appear one (i.e.~singletons) (7), eight
(3), four (3), or ten (6) or fewer times; these pipelines also vary in
whether the minimum abundance threshold should be applied to individual
samples (6, 7) or the pool of samples in a study (3). A notable
exception, the mothur-based pipeline discourages the practice of
removing rare sequences (2). After assigning sequences to ASVs, ASVs are
often analyzed as a taxonomic unit or clustered to generate operational
taxonomic units (OTUs) or phylotypes.

@@

\textbf{Datasets.} I collected 12 publicly available datasets that used
the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
gene from a variety of environments (Table 1; Figure S1).

After removing poor quality and chimeric ASVs and samples that had
uncharacteristically low number of sequences for the dataset, these
datasets included between 7 and 490 samples. The median number of
sequences for each dataset ranged between 6,477 and 193,464. Strikingly,
aside from the marine and soil datasets, which had a relatively small
number of samples, the difference between the sample with the fewest
sequences and the sample with the most sequences for each dataset varied
by between 7.4 and 96.6-fold.

\textbf{The nature of singletons.} Removal of rare ASVs is commonly
justified as a method of removing ASVs that are artifacts. If such ASVs
are artifacts, then one would expect the number of singleton ASVs to
accumulate with sequencing depth. Contrary to this expectation, the
median percentage of sequences that were discarded when singleton ASVs
were removed from each dataset varied between 0.42 and 22.20\%
(bioethanol and seagrass). In addition, with the exception of the
samples from the marine and sediment datasets whose Spearman correlation
was not significantly different from 0.00 (P\textgreater0.05), the
fraction of singleton ASVs in samples was negatively correlated with the
number of sequences in each sample with a range between -0.26 and -0.87
(rice and bioethanol) (Figures 1A and S2). This showed that with
additional sequencing, the probability of seeing singleton ASVs in
multiple samples was greater than the probability of generating an
artifact. This suggests that the singleton ASVs are not as likely to be
artifacts as previously thought. Furthermore, if singleton ASVs were
artifacts, then one would not expect to find them in other samples from
the same dataset. In fact, singleton ASVs from samples with fewer
sequences were often found in samples with more sequences. At least 50\%
of the singleton ASVs found in the samples from the mice, rice,
seagrass, and stream datasets were found in another sample from the same
dataset (Figure 1B). Considering the likelihood of finding an ASV
duplicated in another sample is confounded by the number of samples and
inter-sample diversity, the high coverage of singleton ASVs in these
datasets was remarkable. The correlation between the number of sequences
in a sample and the fraction of that sample's singleton ASVs that were
covered by another sample in the dataset was significant and negative
for 9 of the datasets ranging between -0.33 and -0.84 for the rice and
seagrass datasets, respectively ((Figures 1C and S3)). The negative
correlation indicated that the singleton ASVs in the smaller samples
were more likely to be covered by ASVs in the larger samples. Among the
three datasets without a significant correlation (Spearman correlation,
P\textgreater0.05), the marine and soil datasets had the fewest samples
in our collection and the stream dataset already had a high level of
coverage regardless of the number of sequences. Contrary to the common
motivation for removing rare ASVs, these results indicate that this
practice disproportionately impacts samples with fewer sequences and
likely removes more non-artifact ASVs than those that are artifacts.

@@

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
samples were regenerated by randomly sampling ASVs from the null
distribution as described above such that each samples had the same
number of sequences as it did in the original study. For the second
treatment group, I randomly selected 10\% of the ASVs in the pooled
study distribution to increase their abundance by 5\%. I randomly
generated 100 simulated sets of treatment groups and samples. I then
tested the ability to detect a difference between the two treatment
groups using alpha and beta diversity metrics. The fraction of
significant tests (i.e., P\textless0.05) was a measurement of the
statistical power to detect the difference between the treatment groups.
When considering the differences in richness and diversity, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7).
Among the remaining datasets, the power to detect a difference in the
richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
between 0.10 and 0.53 (rainforest and stream) to detect a difference in
diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
removed, the power to detect a difference in the richness of ASVs
dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
40.0 and 93.3\% for their diversity(rainforest and soil; Figure 4B). The
effect of removing rare ASVs on the richness of OTUs and their diversity
was similar (Figure S7AB). I used the Bray-Curtis dissimilarity index to
compare the simulated communities within each dataset and calculated the
power to detect differences between the two simulated treatment groups
using the analysis of molecular variance (Figure 4C and S7C). Without
removing rare sequences, the power to detect a difference between the
two simulated treatment groups varied between 0.41 and 1.00 (rainforest
and human, stream, and rice). Aside from the bioethanol, human, and mice
datasets, the power to detect differences dropped by between 6.5 and
64.0\% (soil and rice) when singletons were removed. However, when ASVs
that occurred 10 or fewer times were removed from each sample, the power
to detect differences dropped by 12.0 and 97.2\% (human and peromyscus);
similar results were observed when ASVs were clustered into OTUs.
Removing rare ASVs reduced the ability to detect simulated treatment
effects using metrics commonly used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled individual sequences from the null distribution for
each dataset so that each sample had the same number of sequences as it
did in the original dataset. I then determined the richness and
diversity of ASVs and OTUs in each of the samples and randomly assigned
each sample to one of two treatment groups. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (median: 0.049, minimum: 0.041,
maximum: 0.062; Figure 5A and S8A). Similarly, the average fraction of
significant tests did not meaningfully vary from 0.05 when using
analysis of molecular variance to compare communities using Bray-Curtis
distances (median: 0.05, minimum: 0.047, maximum: 0.055; Figure 5A and
S8A). Second, I again sampled sequences from the null distribution, but
assigned samples to one of two treatment groups based on the number of
sequences in each sample. The samples with fewer than the median number
of sequences for the dataset were assigned to one group and those with
more than the median were assigned to the other. This exaggerated bias
has been observed in comparisons of the lung and oral microbiota because
of the larger number of non-specific amplicons that can be sequenced
from lung samples relative to those in the oral cavity leading to a
significant difference in sequencing depth between treatment groups (9).
When rare sequences were not removed, the fraction of significant tests
did not differ from 5\% for comparing the richness, their diversity, or
Bray-Curtis distances (Figure 5B and S8B). However, when rare taxa of
any frequency were removed, the probability of falsely detecing a
difference as signifiant increased with the definition of rarity (Figure
5B and S8B). Not including the small marine dataset, the average
fraction the average fraction of falsely detecting a difference across
datasets when only singletons were removed was 92.45\%. If there is any
relationship between the number of sequences and the treatment group,
the risk of falsely rejecting the null hypothesis is inflated when
researchers use the strategy of removing rare sequences. The most
conservative approach is to not remove low abundance sequences.

@@

The practice of removing rare sequences from samples seems to be a
response to researchers prioritizing the number of reads and length of
sequences over their quality and needing a mechanism to reduce the
inflated number of spurious OTUs (5). Previous work has shown that
assembly of fully overlapping sequence reads results in the lowest
sequencing error rates (2). The studies highlighted in this analysis
sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq
sequencing platform with their version 2 chemistry. The resulting data
consists of two 250 nt reads that span a region that is about 253 nt
long. In contrast, there has been a movement to sequence longer regions
with similar chemistry resulting in less overlap between the sequencing
reads (10, 11). Alternatively, others have prioritized increasing the
number of sequences per sample by sequencing the V4 region, but with
paired or single reads that are 150 nt long (12, 13). Both practices
result in a significantly higher error rate for the resulting assmebly.
Instead, researchers should prioritize the quality over the quantity and
length of their data. For these reasons, this analysis did not use lower
quality data generated by alternative methods.

The impacts of removing rare sequences on the representation of
communities are caused by the uneven impact of applying a single
abundance threshold across all samples. The number of sequences per
sample in the datasets highlighted in this analysis varied by as much as
100-fold (Table 1; Figure S1). If the range in read coverage for a study
is 100-fold, a singleton in a small sample would have had a comparable
relative abundance as a sequence that appeared 100 times in a more
densely sequenced sample. However, it would have been removed from the
smaller sample and not the larger sample. Thus, applying a single
abundance threshold disproportionately impacted the samples with fewer
sequences. This seems to contradict the purpose of removing rare
sequences since a singleton in the smaller sample is more reliable than
the singleton in the larger sample. A superior approach to removing rare
sequences is to use rarefaction to conrol for uneven sampling.

Even with the best sequencing approaches, PCR artifacts and sequencing
errors persist. This may account for some sequences not being observed
other samples. Although this lack of inter-sample coverage could have
been due to treatment effects and natural variation, it is likely that
some portion of the sequences that were unique to samples were artifacts
or contained errors. Researchers are encouraged to use rarefaction to
control for uneven sampling and the presence of spurious sequences. By
rarefying data to a common number of sequences per sample, the number of
spurious sequences can be controlled. As shown in the data I presented,
which was rarefied to a common number of reads per sample within a
dataset, when rare sequences were not removed the power to detect
differences was the highest and the false discovery rate was the
expected 5\% (Figures 5 and S8).

In addition to considerations of how to control for the presence of
spurious sequences, researchers also need to be mindful of how to
interpret the results of their work. Because every dataset will contain
residual sequences that are spurious, measures of richness and diversity
should be made on a relative basis. For example, pronouncements that
communities from an environment or treatment contain a specific number
of taxa are problematic. Instead, we should limit ourselves to
indicating that samples from one treatment group has more taxa than
another without using absolute values of richness. Furthermore, we must
take caution in interpreting rare taxa. Although the data from the
studies highlighted here suggest that most rare sequences are not
spurious, it is likely that some are. Therefore, researchers must
approach rare sequences with more skepticism than more abundant
sequences. Reseachers should seek out other methods to confirm
inferences that they make about rare sequences. This is a standard that
should be applied regardless of their abundance (14).

How to curate and interpret rare sequences has been a significant
challenge since microbial ecologists transitioned away from Sanger
sequencing of samples (15--17). Although the extent of the ``rare
biosphere'' is still an open question, it is important to appreciate the
importance of rare populations in all communities. Populations can be
numerically rare but ubuiquitous or abundant and limited in their
geographic range. Alternatively, they can be numerically rare but
temporally common or abundant but present infrequently. Removing
sequences from any of these settings will limit our ability to study the
role of such populations or the processes that drive their patchy
distributions that are so common with microbial communities (8).

@@

\textbf{Data curation and analysis.} To ensure the highest possible data
quality, datasets were limited to those where the 500 cycle version 2
MiSeq chemistry was used to sequence the amplicons. The paired 250 nt
reads resulted in near complete 2-fold sequencing coverage of every
nucleotide in the ca. 253 nt-long region. This region and sequencing
platform were selected because previous work has shown that a standard
data analysis pipeline in mothur results in a sequencing error rate
below 0.02\% (2). All sequence data were obtained from the Sequence Read
Archive (SRA) and processed using a standard mothur-based sequencing
pipeline that resulted in ASVs as generated by the pre.cluster algorithm
using a threshold of 2 nt (2, 18). ASVs were assigned to OTUs using a
3\% distance threshold using mothur's cluster function with the
OptiClust algorithm (19). To minimize the effects of uneven sampling
effort, samples were rarefied to the number of sequences in the smallest
sample for each dataset. Because metrics of alpha diversity did not
consistently follow a normal distribution, I used the non-parametric
Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis
distances, the amova function within mothur was used, which implements
the analysis of molecular variance algorithm (20).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-Knight2018}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Knight R}, \textbf{Vrbanac A}, \textbf{Taylor
BC}, \textbf{Aksenov A}, \textbf{Callewaert C}, \textbf{Debelius J},
\textbf{Gonzalez A}, \textbf{Kosciolek T}, \textbf{McCall L-I},
\textbf{McDonald D}, \textbf{Melnik AV}, \textbf{Morton JT},
\textbf{Navas J}, \textbf{Quinn RA}, \textbf{Sanders JG},
\textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Tripathi A},
\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhu Q}, \textbf{Caporaso
JG}, \textbf{Dorrestein PC}. 2018. Best practices for analysing
microbiomes. Nature Reviews Microbiology \textbf{16}:410--422.
doi:\href{https://doi.org/10.1038/s41579-018-0029-9}{10.1038/s41579-018-0029-9}.}

\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
a dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform}.
Applied and environmental microbiology \textbf{79}:5112--5120.}

\leavevmode\vadjust pre{\hypertarget{ref-Edgar2016}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
error-correction for illumina 16S and {ITS} amplicon sequencing.
bio{R}xiv. doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.}

\leavevmode\vadjust pre{\hypertarget{ref-Masella2012}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{\textbf{Masella AP}, \textbf{Bartram AK},
\textbf{Truszkowski JM}, \textbf{Brown DG}, \textbf{Neufeld JD}. 2012.
{PANDAseq}: Paired-end assembler for illumina sequences. {BMC}
Bioinformatics \textbf{13}:31.
doi:\href{https://doi.org/10.1186/1471-2105-13-31}{10.1186/1471-2105-13-31}.}

\leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
vastly improves diversity estimates from illumina amplicon sequencing.
Nature Methods \textbf{10}:57--59.
doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}

\leavevmode\vadjust pre{\hypertarget{ref-Amir2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
\textbf{Navas-Molina JA}, \textbf{Kopylova E}, \textbf{Morton JT},
\textbf{Xu ZZ}, \textbf{Kightley EP}, \textbf{Thompson LR}, \textbf{Hyde
ER}, \textbf{Gonzalez A}, \textbf{Knight R}. 2017. Deblur rapidly
resolves single-nucleotide community sequence patterns. {mSystems}
\textbf{2}.
doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.}
